---
title: "Assignment 5: Data Visualization"
author: "Sayra Martinez"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
#1 Installing libraries, setting my directory and importing my datasets
library(tidyverse);library(lubridate);library(here); library(dplyr); 
library(cowplot); library(viridis); 
library(RColorBrewer); library(colormap)
getwd()

PeterPaul.chem.nutrients <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
           stringsAsFactors = T)

Litter <-
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
           stringsAsFactors = T)

#2 Setting and verifying sampledate & collectDate as date
class(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients$sampledate <- mdy(PeterPaul.chem.nutrients$sampledate)
class(PeterPaul.chem.nutrients$sampledate)

class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate)
class(Litter$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Creating my theme

sayra.theme <- 
  theme(axis.text = element_text(color = "black"),
        
        legend.position = "right",
        legend.direction = "vertical",
        plot.title =        element_text(color = "black", size = 10),
        axis.ticks =        element_line(6),
        axis.text.x = element_text(color = "darkred", size = 7),
        axis.text.y = element_text(color = "darkred", size = 7),
        plot.background = element_rect(fill = "gray90"),
        legend.background = element_rect(
      color="black"
      ),
    legend.title = element_text(
      color="royalblue", size = 7.5
    )
  )

theme_set(sayra.theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

Phosphorus_Phosphate <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = po4, color = depth)) +
  geom_point(alpha=0.5, size= 1) +
  xlim(0, 100) +
  ylim(0, 20)+ 
  facet_wrap(vars(lakename)) +
  ggtitle("Phosphorus and Phosphate") +
  labs(color='Depth') + 
  xlab(expression(paste("Total Phosphate"))) +
  ylab(expression(paste("Phosphate"))) +
  geom_smooth(method = lm)
print(Phosphorus_Phosphate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 I create the plotboxes, but first I put month as a factor

PeterPaul.chem.nutrients$month_f <- factor(
  PeterPaul.chem.nutrients$month, 
  levels=1:12,
  labels=month.abb)

Temp.Boxplot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month_f, y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(drop=F) +
  ylab(
    expression(paste("C°")))+
  theme(axis.text.x = element_blank(),         
        legend.position =  "none",
         axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7))
print(Temp.Boxplot)

Phospho.boxplot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month_f, y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(drop=F) +
  ylab(
    expression(paste(
    "Total Phosphorus")))+
  theme(axis.text.x = element_blank(),         
        legend.position =  "none",
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 7))
print(Phospho.boxplot)

Nitro.boxplot <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month_f, y = tn_ug)) +
  geom_boxplot(
    aes(color = lakename)) +
  scale_x_discrete(drop=F) +
  labs(color = "Lake name")+
  xlab(expression(paste("Month")))+
  ylab(expression(paste(
    "Total Nitrogen")))+
  theme(legend.position =  "bottom",
        legend.direction =  "horizontal",
  axis.title.x = element_text(size = 7),
  axis.title.y = element_text(size = 7),
  legend.key.size = unit(0.5, "line"))
print(Nitro.boxplot)

Three.Nutrients <-
plot_grid(Temp.Boxplot, Phospho.boxplot, Nitro.boxplot, nrow = 3, rel_heights = c(1, 1.5, 2.5), align = T)
print(Three.Nutrients)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: In both cases, as water's temperature increases, during the summer (from June to August), the TP and TN levels are also higher, specially in the case of Peter Lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6 Create subgroup and plot, using NLCD classes as colors

Litter.Needles <- filter(Litter,
                functionalGroup
         == "Needles")

Needles.plot <-  Litter.Needles %>%
  filter(functionalGroup
         == "Needles") %>%
ggplot(aes(x = collectDate, 
          y = dryMass, color = nlcdClass)) +
  geom_point() +
  labs(title = 
         "Dry Mass. Needle Litter Functional Group")+
  theme(legend.position =  "right",
        legend.direction =  "vertical")+
  labs(color = "NLCD classes")
Needles.plot


#7 Plot with NLCD separated

Litter.Needles <- filter(Litter,
                functionalGroup
         == "Needles")

Needles.plot2 <-  Litter.Needles %>%
  filter(functionalGroup
         == "Needles") %>%
ggplot(aes(x = collectDate, 
          y = dryMass, color = nlcdClass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass)) +
  labs(title = "Dry Mass. Needle Litter Functional Group")+
  theme(legend.position =  "right",
        legend.direction =  "vertical")
Needles.plot2

# I put them together to answer
Needles.Together <-
plot_grid(Needles.plot, Needles.plot2, nrow = 2, rel_heights = c(1.2, 1.2))
print(Needles.Together)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: The one separating NLCD classes is more effective in this case, because it allows to observe clearly the Dry Mass of each of them;  while in the one that only put them as colors, they overlap and it is no clear the observations.