---
title: "Assignment 10: Data Scraping"
author: "Sayra Martinez"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 Loading packages and checking directory
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
library(tidyverse);library(rvest); library(dataRetrieval);library(tidycensus)
library(ggplot2); library(scales); library("ggplot2"); library("lubridate"); library("dplyr")
getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
URL <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022'
water_website <- read_html(URL)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
# From System information section
system_name <- water_website %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text(trim=T)
PWSID <- water_website %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text(trim=T)
ownership <- water_website %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text(trim=T)
# From water supply sources
MGD <- water_website %>% html_nodes("th~ td+ td") %>% html_text(trim=T) %>% as.numeric()

# Just to check the extracted data
print(list(
  System_Name = system_name,
  PWSID = PWSID,
  Ownership = ownership,
  MGD = MGD
))

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 Preparing the information
System_name <- rep("Durham", 12)
PWSID <- rep("03-32-010", 12)
Ownership <- rep("Municipality", 12)
Month <- water_website %>% html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% html_text(trim=T)
Year <- rep("2022",12)
Date <- as.Date(paste("2022", Month, "01", sep = "-"), format = "%Y-%b-%d")

#Creating my data frame
Water_Supply <- data.frame(
  System_Name = System_name,
  PWSID = PWSID,
  Ownership = Ownership,
  MGD = MGD,
  Date = Date,
  Month = Month)

#5 
Water_Durhamplot <- ggplot(Water_Supply, aes(x = Date, y = MGD)) +
  geom_line(group = 1) +
  labs(x = "Month in 2022", y = "Maximum Daily Withdrawals") +
  theme_linedraw() +
  ggtitle("Maximum Daily Withdrawals - Monthly Data for 2022") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")
Water_Durhamplot

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scraping <- function(PWSID, Year) {
  the_URL <- paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', PWSID, '&year=', Year)
  water_website <- read_html(the_URL)

#scraping data from the given website  
system_name <- water_website %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
PWSID <- water_website %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text(trim=T)
ownership <- water_website %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text(trim=T)
# From water supply sources
MGD <- water_website %>% html_nodes("th~ td+ td") %>% html_text(trim=T) %>% as.numeric()

#Creating Date information
#Month <- water_website %>% html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% html_text(trim=T) #Although it was correctly extracting the information for Durham, it didn't work with the rest, so I decided to do it manually.
Month <- c("Jan","May","Sep","Feb","Jun","Oct","Mar","Jul","Nov","Apr","Aug","Dec")
Date <- as.Date(paste(Year, Month, "01", sep = "-"), format = "%Y-%b-%d")

#Into data frame (I tried to make the the rep and creating the Dates directly and yes, it is possible)
Water_Supply_scraping <- data.frame(
  System_Name = rep(system_name, 12),
  PWSID = rep(PWSID, 12),
  Ownership = rep(ownership, 12),
  MGD = MGD,
  Year = rep(Year, 12),
  Date = Date)
Water_Supply_scraping 
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
Durham_2015 <- scraping('03-32-010','2015')
Durham_2015

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
Asheville_2015 <- scraping('01-11-010','2015')
Asheville_2015

two_towns <- bind_rows(Durham_2015,Asheville_2015)

ggplot(two_towns, aes(x = Date, y = MGD, color = System_Name)) +
  geom_line() +
  labs(x = "Month", y = "Maximum Daily Withdrawals", color = "Town") +
  theme_light() +
  ggtitle("Maximum Daily Withdrawals in Asheville and Durham. 2015") +
  theme(plot.title = element_text(hjust = 0.1))+
  scale_x_date(date_labels = "%b", date_breaks = "1 month")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
Asheville_decade <- 2010:2021
Asheville_id <- rep('01-11-010', length(Asheville_decade))

Asheville_df <- map2(Asheville_id, Asheville_decade, scraping)
Asheville_df <- bind_rows(Asheville_df)

Asheville_plot <- ggplot(Asheville_df, aes(x = Date, y = MGD)) +
  geom_line() +
  geom_smooth(method = 'loess', se = F, color = 'blue') +
  labs(x = "Date", y = "Maximum Daily Withdrawals", color = "Town") +
  theme_light() +
  ggtitle("Maximum Daily Withdrawals in Asheville: 2010-2021")
Asheville_plot
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Yes, the plot shows that from 2010 and around 2015, the water usage was slighltly decreasing, but it recovers and started increasing from then. 
>
